---
title: "Exploratory Data Analysis"
author: "Yijia Jiang"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r, message=FALSE, include=FALSE}
library(tidyverse)
library(survival)
library(table1)
library(kableExtra)
library(ggplot2)
library(GGally)
```


# 1. Data Overview
In this study, the dataset we used is called ‘colon’, which is found in the ‘survival’ R package. These are data from one of the first successful trials of adjuvant chemotherapy for colon cancer. Levamisole is a low-toxicity compound previously used to treat worm infestations in animals; 5-FU is a moderately toxic (as these things go) chemotherapy agent. There are two records per person, one for recurrence and one for death. There are 1858 observations and 16 variables. 

* id: patient id 
* study: 1 for all patients 
* rx: treatment - Obs(ervation), Lev(amisole), Lev(amisole)+5-FU 
* sex: sex (0 = Female, 1 = Male)
* age: observation age in years 
* obstruct: obstruction of colon by tumour (0 = No, 1 = Yes)
* perfor: perforation of colon (0 = No, 1 = Yes)
* adhere: adherence to nearby organs (0 = No, 1 = Yes)
* nodes: number of lymph nodes with detectable cancer 
* time: days until event or censoring 
* status: censoring status (0 = Censored, 1 = Event)
* differ: differentiation of tumour (1 = Well, 2 = Moderate, 3 = Poor) 
* extent: extent of local spread (1 = Submucosa, 2 = Muscle, 3 = Serosa, 4 = Contiguous structures) 
* surg: time from surgery to registration (0 = Short, 1 = Long) 
* node4: more than 4 positive lymph nodes (0 = No, 1 = Yes)
* etype: event type (1 = recurrence, 2 = death)


The primary endpoints are the death of patients and the recurrence of patients. The type of censoring is right censoring, which means patients left the study before their death. 

```{r message = FALSE, warning = FALSE}
data(cancer, package = "survival") 
colon_tb <- as_tibble(colon)

colon_tb$sex <- factor(colon_tb$sex, levels = c(0,1), labels = c("Female", "Male"))
colon_tb$obstruct <- factor(colon_tb$obstruct, levels = c(0,1), labels = c("No", "Yes"))
colon_tb$perfor <- factor(colon_tb$perfor, levels = c(0,1), labels = c("No", "Yes"))
colon_tb$adhere <- factor(colon_tb$adhere, levels = c(0,1), labels = c("No", "Yes"))
colon_tb$status <- factor(colon_tb$status, levels = c(0,1), labels = c("Censored", "Dead"))
colon_tb$differ <- factor(colon_tb$differ, levels = c(1,2,3), labels = c("Well", "Moderate","Poor"))
colon_tb$extent <- factor(colon_tb$extent, levels = c(1,2,3,4), labels = c("Submucosa", "Muscle","Serosa","Contiguous structures"))
colon_tb$surg <- factor(colon_tb$surg, levels = c(0,1), labels = c("Short", "Long"))
colon_tb$node4 <- factor(colon_tb$node4, levels = c(0,1), labels = c("No", "Yes"))

label(colon_tb$rx) <- "Treatment"
label(colon_tb$sex) <- "Sex"
label(colon_tb$age) <- "Age"
label(colon_tb$obstruct) <- "Obstruction of colon by tumour"
label(colon_tb$perfor) <- "Perforation of colon"
label(colon_tb$adhere) <- "Adherence to nearby organs"
label(colon_tb$nodes) <- "Number of lymph nodes with detectable cancer"
label(colon_tb$time) <- "Time until event or censoring"
label(colon_tb$status) <- "Censoring status"
label(colon_tb$differ) <- "Differentiation of tumour"
label(colon_tb$extent) <- "Extent of local spread"
label(colon_tb$surg) <- "Time from surgery to registration"
label(colon_tb$node4) <- "More than 4 positive lymph nodes"

units(colon_tb$age) <- "years"
units(colon_tb$time) <- "days"

table1(~ rx + sex + age + obstruct + perfor + adhere + differ + extent + surg + nodes + node4| status, data = colon_tb %>% subset(etype == 1), overall = "Total")
table1(~ rx + sex + age + obstruct + perfor + adhere + differ + extent + surg + nodes + node4| status, data = colon_tb %>% subset(etype == 2), overall = "Total")
```

# 2. Corrlation between Variables
```{r}
pairs(~ status + rx + sex + age + obstruct + perfor + adhere + differ + extent + surg + nodes + node4, data = colon %>% subset(etype == 1), panel = panel.smooth, upper.panel = NULL, main = "Scatterplot Matrix")

pairs(~ status + rx + sex + age + obstruct + perfor + adhere + differ + extent + surg + nodes + node4, data = colon %>% subset(etype == 2), panel = panel.smooth, upper.panel = NULL, main = "Scatterplot Matrix")

colon %>% 
  subset(etype == 1) %>% 
  mutate(rx = as.numeric(rx)) %>% 
  dplyr::select(-id, -study,-etype, -time) %>%  
  ggcorr(label = TRUE, hjust = 0.9, layout.exp = 2, label_size = 3, label_round = 2)

colon %>% 
  subset(etype == 2) %>% 
  mutate(rx = as.numeric(rx)) %>% 
  dplyr::select(-id, -study,-etype, -time) %>%  
  ggcorr(label = TRUE, hjust = 0.9, layout.exp = 2, label_size = 3, label_round = 2)
```


