---
title: "Survival Analysis of Mortality of Adjuvant Chemotherapy for Colon Cancer"
author: "Yijia Jiang, Ziyan Xu"
date: "`r Sys.Date()`"
output: pdf_document
---


```{r, message=FALSE, include=FALSE}
library("tidyverse")
library("survival")
library("ggplot2")
library("survminer")
library("MASS")
library("gridExtra")
```


# 1. Data Tidy 
Recall that there are two records for each patient indicated by the event type (etype) variable, where etype == 1 refers to the event of a recurrence and etype == 2 indicates death. In order to answer our first research question, which is to study the time until death, we must create a marginal model by subsetting the colon data to only include the event of mortality. 

```{r, message=FALSE, warning=FALSE}
# import the dataset from survival package
colon <- survival::colon

# tidy mortality dataset
colon.death <- colon %>% filter(etype == '2') %>% 
               dplyr::select(-c(id,study,etype, node4)) %>% 
               drop_na() %>% 
               mutate(rx = as.factor(rx))
```

&nbsp;


# 2. Kaplan-Meier Survival Estimate

```{r, message=FALSE, warning=FALSE}
death.fit <- survfit(Surv(time,status) ~ rx, data = colon.death)
options(max.print = 10000)
print(summary(death.fit))
ggsurvplot(death.fit, conf.int = F, break.time.by = 500, pval = TRUE,
           font.x.size = 12, font.y.size = 12, font.legend.size = 9, surv.median.line = "hv",
           legend.title = "Treatment", legend.labs = c("Observation", "Levamisole","Levamisole+5-FU"),
           title = "Kaplan-Meier Curve for Colon Cancer Mortality \nby Treatment", 
           xlab = "Time (until death) in Days", 
           risk.table = T, risk.table.height = 0.25, risk.table.fontsize = 4, 
           tables.theme = theme_cleantable())
```

From the plot above, there is some indication that patients who received the adjuvant treatment with levamisole plus fluorouracil (Lev + 5FU) have a higher survival probability than patients with no further treatment and patients who received the treatment with levamisole alone. The median survival time for observation group and levamisole group are approximately 2100 days and 2200 days, respectively. However, until the end of the trial, the survival probability of Levamisole + 5-FU treatment group is greater than 50% as we fail to observe the curve crossing the 50% line.

&nbsp;


# 3. Log-Rank Test

Noticing the difference of survival probability between the three treatment groups, we conduct a Log-rank hypothesis test to test the null hypothesis of no difference among the three treatments in the mortality model.
```{r, message=FALSE, warning=FALSE}
survdiff(Surv(time, status) ~ rx, data = colon.death)
```

<<<<<<< HEAD
From this log-rank test, we get a p-value that is closed to 0.004, which is significant at a 0.05 level. We want to conclude that there is a significant difference among the three treatments in the mortality model.
=======
From this log-rank test, we get a p-value equal to 0.005, which indicates a significant difference among treatment groups at a 0.05 level. 
>>>>>>> fc812e0f2fdf362b59d99fe5017fcf88363010b4

&nbsp;


# 4. Cox PH Model 
## 4.1 Model Selection

We now use automatic stepwise selection with Akaike information criterion (AIC) to determine the covariates that best represent an appropriate cox proportional hazards model for the event of death. 

```{r, message=FALSE, warning=FALSE}
# fit all the variables in the model
d.model.full <- coxph(Surv(time, status) ~ ., data = colon.death)

# stepwise selection with AIC criterion
d.model.aic <- step(d.model.full, direction = "both", k = 2)
```

<<<<<<< HEAD
The resulting model with the lowest AIC is:
Surv(time, status) ~ obstruct + differ + extent + surg + nodes + age + rx
=======
The resulting model with the AIC is:
Surv(time, status) ~ obstruct + differ + extent + surg + node4 + age + rx
>>>>>>> fc812e0f2fdf362b59d99fe5017fcf88363010b4


Next, we used the Analysis of Deviance procedure to get the proper Likelihood Ratio Test to confirm if each of the covariates selected by the stepwise selection method is significant to include in the Cox Proportional Model.

```{r, message=FALSE, warning=FALSE}
anova(d.model.aic)
```

We can see that p-values for covariates "obstruct", "differ", "extent", "surg", "nodes" and "rx" are much smaller than 0.05, except for "age", indicating that they have a significant effect on time until death. Therefore, we will include these 6 covariates in our Cox PH model.

Therefore, we obtain the resulting model, which is Surv(time, status) ~ obstruct + differ + extent + surg + nodes + rx.

&nbsp;


## 4.2 Model Diagnostics 
### 4.2.1 Check proportionality of hazard ratios

**Log of Negative Log of Estimated Survival Function** 

To check the proportional hazards assumption for this model, we make diagnostic plots using log of negative log of estimated survival function. First comes to covariate obstruct.

```{r, message=FALSE, warning=FALSE}
d.obstruct.fit <- survfit(Surv(time, status) ~ obstruct, data = colon.death)
cloglog_obstruct <- ggsurvplot(d.obstruct.fit, conf.int = F, font.x.size = 12, font.y.size = 12, font.legend.size = 9,
                               fun = "cloglog",
                               xlim = c(20, 4000),
                               xlab = "Time (until death) in Days)",
                               title = "Log of Negative Log of Estimated Survival Function \nfor Colon Cancer Mortality by obstruct",
                               legend.title = "Obstruction of colon by tumour",
                               legend.labs = c("0-No", "1-Yes"))

cloglog_obstruct
```

According to the plot, the distance between two obstruct curves begin to narrow after 2000 days which causes some concern that the assumption might be violated. However, it is also reasonable to assume that the curves are wider apparent earlier in the study since there are less occurrences of death before 2000 days. Hence we believe it is best to ignore the noisiness of the plot since the curves are roughly parallel after 2000 days. Thus, the cox proportional hazards assumption is valid to use for "obstruct".


We continue to plot the C-log-log plot for the covariate differ.

```{r, message=FALSE, warning=FALSE}
d.differ.fit <- survfit(Surv(time, status) ~ differ, data = colon.death)
cloglog_differ <- ggsurvplot(d.differ.fit, conf.int = F, font.x.size = 12, font.y.size = 12, font.legend.size = 9,
                             fun = "cloglog",
                             xlim = c(20, 4000),
                             xlab = "Time (until death) in Days)",
                             title = "Log of Negative Log of Estimated Survival Function \nfor Colon Cancer Mortality by differ",
                             legend.title = "Differentiation of tumour",
                             legend.labs = c("1-Well", "2-Moderate","3-Poor"))

cloglog_differ
```

According to the plot, the curves in the C-log-log plot are crossing over after 300 days. The assumption of proportional hazard ratio among the three differ groups is violated as we fail to see three parallel lines against log time.


We continue to plot the C-log-log plot for the covariate extent.

```{r, message=FALSE, warning=FALSE}
d.extent.fit <- survfit(Surv(time, status) ~ extent, data = colon.death)
cloglog_extent <- ggsurvplot(d.extent.fit, conf.int = F, font.x.size = 12, font.y.size = 12, font.legend.size = 9,
                             fun = "cloglog",
                             xlim = c(20, 4000),
                             xlab = "Time (until death) in Days)",
                             title = "Log of Negative Log of Estimated Survival Function \nfor Colon Cancer Mortality by extent",
                             legend.title = "Extent of local spread",
                             legend.labs = c("1-Submucosa", "2-Muscle", "3-Serosa", "4-Contiguous structures"))

cloglog_extent
```

According to the plot, the curves in the C-log-log plot are crossing over after 100 days. The assumption of proportional hazard ratio of extent is violated as we fail to see four parallel lines against log time.


We continue to plot the C-log-log plot for the covariate surg.

```{r, message=FALSE, warning=FALSE}
d.surg.fit <- survfit(Surv(time, status) ~ surg, data = colon.death)
cloglog_surg <- ggsurvplot(d.surg.fit, conf.int = F, font.x.size = 12, font.y.size = 12, font.legend.size = 9,
                           fun = "cloglog",
                           xlim = c(20, 4000),
                           xlab = "Time (until death) in Days)",
                           title = "Log of Negative Log of Estimated Survival Function \nfor Colon Cancer Mortality by surg",
                           legend.title = "Time from surgery to registration",
                           legend.labs = c("0-Short", "1-Long"))

cloglog_surg
```

According to the plot, the distance between two surg curves begin to narrow after 100 days and becomes parallel to each other. We can assume that the cox proportional hazards assumption is valid to use for `surg`.



We continue to plot the C-log-log plot for the covariate rx.

```{r, message=FALSE, warning=FALSE}
d.surg.fit <- survfit(Surv(time, status) ~ rx, data = colon.death)
cloglog_rx <- ggsurvplot(d.surg.fit, conf.int = F, font.x.size = 12, font.y.size = 12, font.legend.size = 9,
                         fun = "cloglog",
                         xlim = c(20, 4000),
                         xlab = "Time (until death) in Days)",
                         title = "Log of Negative Log of Estimated Survival Function \nfor Colon Cancer Mortality by rx",
                         legend.title = "Treatment",
                         legend.labs = c("Observation", "Levamisole","Levamisole+5-FU"))

cloglog_rx
```

According to the plot, the distance between three treatment curves begin to narrow after 2000 days which causes some concern that the assumption might be violated. However, it is also reasonable to assume that the curves are wider apparent earlier in the study since there are less occurrences of death before 2000 days. Hence we believe it is best to ignore the noisiness of the plot since the curves are roughly parallel after 2000 days. Thus, the cox proportional hazards assumption is valid to use for this covariate.



```{r, message=FALSE, warning=FALSE}
# combine all the cloglog plots and save them to a pdf file
splots <- list()
splots[[1]] <- cloglog_obstruct 
splots[[2]] <- cloglog_differ
splots[[3]] <- cloglog_extent
splots[[4]] <- cloglog_surg
splots[[5]] <- cloglog_rx
cloglog_plot = arrange_ggsurvplots(splots, print = FALSE, ncol = 2, nrow = 3)

ggsave(cloglog_plot,file = "./plot/d.C-log-log-plots.pdf",width = 12,height = 15)
ggsave(cloglog_plot,file = "./plot/d.C-log-log-plots.png",width = 12,height = 15)
```

&nbsp;  


**Schoenfeld residuals**

```{r, message=FALSE, warning=FALSE}
czph <- cox.zph(coxph(Surv(time, status) ~ obstruct + differ + extent + surg + nodes + rx, data = colon.death))
czph
schoenfeld_plot <- ggcoxzph(czph, font.main = 10, font.x = 10, font.y = 10, font.tickslab = 8,
                            point.alpha = 0.5, point.col = "grey25")

schoenfeld_plot
ggsave("./plot/d.schoenfeld residual_plots.pdf", arrangeGrob(grobs = schoenfeld_plot))
ggsave("./plot/d.schoenfeld residual_plots.png", arrangeGrob(grobs = schoenfeld_plot))
```
From the output above, the tests for covariates "obstruct", "differ" are statistically significant, and the global test is also statistically significant. Therefore, we can assume the violation of proportional hazards on these covariates, which requires corrections of non-proportional hazard ratio.



### 4.2.2 Test influential observations 

```{r, message=FALSE, warning=FALSE}
# check outliers in term of dfbeta
outlier_dfbeta <- ggcoxdiagnostics(coxph(Surv(time, status) ~ obstruct + differ + extent + surg + nodes + rx, data = colon.death), 
                                   type = "dfbeta", linear.predictions = FALSE, ggtheme = theme_bw())
outlier_dfbeta
ggsave(outlier_dfbeta, file = "./plot/d.outlier_dfbeta.pdf")
ggsave(outlier_dfbeta, file = "./plot/d.outlier_dfbeta.png")

outlier_deviance <- ggcoxdiagnostics(coxph(Surv(time, status) ~ obstruct + differ + extent + surg + nodes + rx, data = colon.death), 
                                   type = "deviance", linear.predictions = FALSE, ggtheme = theme_bw())

outlier_deviance
```
It’s also possible to check outliers by visualizing the influence of each point, in terms of DFBETA - the impact on the coefficient of covariates in the model were that specific point to be removed from the data set. In general, we can identify the observations with positive and negative DFBETAs. Comparing the magnitudes of the largest dfbeta values to the regression coefficients suggests that none of the observations is terribly influential individually, even though some of the dfbeta values for extent are large compared with the others.


## 4.3 Corrections for violation of the PH Assumption
```{r, message=FALSE, warning=FALSE}
d.model.inter = coxph(Surv(time, status) ~ rx + obstruct + differ + extent + surg + nodes + tt(obstruct) + tt(differ) , data = colon.death, tt = function(x,t, ...) x*log(t))

summary(d.model.inter)
anova(d.model.inter)
```

We observe significance from Schoenfeld residuals, so we modify the model by adding the interaction of covariates with function of time.

&nbsp;


## 4.4 Final model

With the inclusion of treatment, our final model is given by: Surv(time, status) ~ rx + obstruct + differ + extent + surg + nodes + tt(obstruct) + tt(differ) , data = colon.death, tt = function(x,t, ...) x*log(t)).

```{r, message=FALSE, warning=FALSE}
d.model.inter = coxph(Surv(time, status) ~ rx + obstruct + differ + extent + surg + nodes + tt(obstruct) + tt(differ) , data = colon.death, tt = function(x,t, ...) x*log(t))

summary(d.model.inter)
```

&nbsp;

# 5. Discussion

Our final model is given by: Surv(time, status) ~ rx + obstruct + differ + extent + surg + nodes + tt(obstruct) + tt(differ).

We can now observe whether treatments helps improve the survival rate in colon cancer patients from negative coefficients of different treatment groups. 

Our result indicates that the hazard ratio for the group treated with rxLev is 0.934, corresponding to a 6.6% decrease in risk of death compared to the observation group. Since the p-value for the coefficient is 0.552 which is larger than 0.05, there is not significant evidence to indicate a survival difference between the group treated with rxLev and the observation group. Additionally, we are 95% confident that the true risk of death is between [0.747, 1.169]. Therefore, we conclude that the treatment rxLev is not effective on improving the survival probability in colon cancer patients.


On the other hand, our result indicates that the hazard ratio for the group treated with rxLev+5FU is 0.700, corresponding to a 30% decrease in risk of death compared to the observation group. Since the p-value for the coefficient is 0.003 which is samller than 0.05, there is significant evidence to indicate a survival difference between the group treated with rxLev+5FU and the observation group. Additionally, we are 95% confident that the true risk of death is between [0.552, 0.889]. Therefore, we conclude that the treatment rxLev+5FU is effective on improving the survival probability in colon cancer patients.

Therefore, we conclude that the treatment Levamisole+5-FU is effective as an adjuvant Chemotherapy on improving the survival rate in colon cancer patients.

&nbsp;


 



 
