---
title: "Survival Analysis of Recurrence of Adjuvant Chemotherapy for Colon Cancer"
author: "Kaitlyn Wang"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r}
library(survival)
library(tidyverse)
library(ggplot2)
library(survminer)
library(MASS)
library(rms)
library(gridExtra)
```

# 1. Data Import and Data Cleaning
```{r}
colon <- survival::colon
colon.recur <- colon %>% filter(etype == '1') %>% 
               dplyr::select(-c(nodes,id,study,etype)) %>% 
               mutate(rx = as.factor(as.numeric(rx)),
                      ) %>% 
               drop_na()
```

&nbsp;

# 2. Kaplan-Meier Survival Estimate
```{r}
recur.fit <- survfit(Surv(time, status) ~ rx, data = colon.recur)

ggsurvplot(recur.fit, conf.int = F, break.time.by = 500,
           font.x.size = 12, font.y.size = 12, font.legend.size = 9, surv.median.line = "hv", 
           legend.title = "Treatment", legend.labs = c("Observation", "Levamisole","Levamisole+5-FU"),
           title = "Kaplan-Meier Curve for Colon Cancer Recurrence \nby Treatment", 
           xlab = "Time in Days", 
           risk.table = T, risk.table.height = 0.25, risk.table.fontsize = 4, 
           tables.theme = theme_cleantable())
```

&nbsp;

# 3. Log-Rank Test
```{r}
coxph(Surv(time, status) ~ rx, data = colon.recur)
summary(coxph(Surv(time, status) ~ rx, data = colon.recur))
```

&nbsp;

# 4. Cox PH Model 
## 4.1 Model Selection
```{r}
# full model
r.model.full <- coxph(Surv(time, status) ~ ., data = colon.recur)

# stepwise selection with AIC criterion
r.model.aic <- step(r.model.full, direction = "both", k = 2)

anova(r.model.aic)
```

Stepwise selection method selected model:
Surv(time, status) ~ differ + rx + extent + surg + node4

&nbsp;

## 4.2 Model Diagnostic 
### 4.2.1 Check proportionality of hazard ratios
**Log of Negative Log of Estimated Survival Function**
```{r}
r.differ.fit <- survfit(Surv(time, status) ~ differ, data = colon.recur)
cll.differ = ggsurvplot(r.differ.fit, conf.int = F, font.x.size = 12, font.y.size = 12, font.legend.size = 9,
           fun = "cloglog",
           xlim = c(20, 4000),
           xlab = "Time (Days)",
           legend.lab = c("1-well","2-moderate","3-poor"),
           legend.title = 'Differ',
           title = "Log of Negative Log of Estimated Survival Function \nfor Colon Cancer Recurrence by Differ")
```

```{r}
r.extent.fit <- survfit(Surv(time, status) ~ extent, data = colon.recur)
cll.extent = ggsurvplot(r.extent.fit, conf.int = F, font.x.size = 12, font.y.size = 12, font.legend.size = 9,
           fun = "cloglog",
           xlim = c(20, 4000),
           xlab = "Time (Days)",
           legend.lab = c("1-submucosa","2-muscle","3-serosa", "4-contiguous structures"),
           legend.title = 'Extent',
           title = "Log of Negative Log of Estimated Survival Function \nfor Colon Cancer Recurrence by Extent")
```

```{r}
r.surg.fit <- survfit(Surv(time, status) ~ surg, data = colon.recur)
cll.surg = ggsurvplot(r.surg.fit, conf.int = F, font.x.size = 12, font.y.size = 12, font.legend.size = 9,
           fun = "cloglog",
           xlim = c(20, 4000),
           xlab = "Time (Days)",
           legend.lab = c("0-short","1-long"),
           legend.title = 'Surg',
           title = "Log of Negative Log of Estimated Survival Function \nfor Colon Cancer Recurrence by Surg")
```

```{r}
r.node4.fit <- survfit(Surv(time, status) ~ node4, data = colon.recur)
cll.node4 = ggsurvplot(r.node4.fit, conf.int = F, font.x.size = 12, font.y.size = 12, font.legend.size = 9,
           fun = "cloglog",
           xlim = c(20, 4000),
           xlab = "Time (Days)",
           legend.lab = c("0 = No", "1 =Yes"),
           legend.title = 'node4',
           title = "Log of Negative Log of Estimated Survival Function \nfor Colon Cancer Recurrence by Nodes")
```

```{r}
splots <- list()
splots[[1]] <- cll.differ
splots[[2]] <- cll.extent
splots[[3]] <- cll.surg
splots[[4]] <- cll.node4

cloglog_plot = arrange_ggsurvplots(splots, print = FALSE, ncol = 2, nrow = 2)
ggsave(cloglog_plot,file = "./plot/r.C-log-log-plots.pdf",width = 12,height = 15)
ggsave(cloglog_plot,file = "./plot/r.C-log-log-plots.png",width = 12,height = 15)

```

**Schoenfeld residuals**
```{r}
r.residual = cox.zph(coxph(Surv(time, status) ~ differ + rx + extent + surg + node4, data = colon.recur))
r.residual
residual_plot = ggcoxzph(r.residual, font.main = 10, font.x = 10, font.y = 10, font.tickslab = 8,
         point.alpha = 0.5, point.col = "grey25")

ggsave("./plot/d.schoenfeld residual_plots.pdf", arrangeGrob(grobs = residual_plot))
ggsave("./plot/d.schoenfeld residual_plots.png", arrangeGrob(grobs = residual_plot))
```
differ and node4 violates ph assumption

&nbsp;


## 4.3 Modification for Violation of PH Assumption
```{r}
#add interaction of covariate with function of time
colon.recur1 = colon.recur %>% 
               mutate(differ_time = as.numeric(differ)*log(time),
                      node4_time = as.numeric(node4)*log(time))

r.model.inter = coxph(Surv(time, status) ~ rx + extent + surg + differ + node4 + differ_time + node4_time, data = colon.recur1)
anova(r.model.inter)
```

&nbsp;

## 4.4 Final model
```{r}
r.final.model = coxph(Surv(time, status) ~ rx + extent + surg + differ + node4 + differ_time + node4_time, data = colon.recur1)
summary(r.final.model)
```

```{r}
covariates <- c("rx", "extent","surg","differ","node4","differ_time","node4_time")
univ_formulas <- sapply(covariates,
                        function(x) as.formula(paste('Surv(time, status)~', x)))
                        
univ_models <- lapply( univ_formulas, function(x){coxph(x, data = colon.recur1)})
# Extract data 
univ_results <- lapply(univ_models,
                       function(x){ 
                          x <- summary(x)
                          p.value<-signif(x$wald["pvalue"], digits=2)
                          wald.test<-signif(x$wald["test"], digits=2)
                          beta<-signif(x$coef[1], digits=2);#coeficient beta
                          HR <-signif(x$coef[2], digits=2);#exp(beta)
                          HR.confint.lower <- signif(x$conf.int[,"lower .95"], 2)
                          HR.confint.upper <- signif(x$conf.int[,"upper .95"],2)
                          HR <- paste0(HR, " (", 
                                       HR.confint.lower, "-", HR.confint.upper, ")")
                          res<-c(beta, HR, wald.test, p.value)
                          names(res)<-c("beta", "HR (95% CI for HR)", "wald.test", 
                                        "p.value")
                          return(res)
                          #return(exp(cbind(coef(x),confint(x))))
                         })
res <- t(as.data.frame(univ_results, check.names = FALSE))
as.data.frame(res)
```

