---
title: "survival project"
author: "Kaitlyn Wang"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(survival)
library(tidyverse)
library(ggplot2)
library(survminer)
library(rms)
```

# 1. Data Import and Data Cleaning
```{r}
colon <- survival::colon
colon.recur <- colon %>% filter(etype == '1') %>% 
               select(-c(id,study,etype, node4)) %>% 
               drop_na() %>% 
               mutate(rx = as.factor(rx))
```

# 2. Kaplan-Meier Survival Estimate
```{r}
recur.fit <- survfit(Surv(time, status) ~ rx, data = colon.recur)

ggsurvplot(recur.fit, conf.int = F, break.time.by = 500, pval = TRUE,
           font.x.size = 12, font.y.size = 12, font.legend.size = 9, surv.median.line = "hv", 
           legend.title = "Treatment", legend.labs = c("Observation", "Levamisole","Levamisole+5-FU"),
           title = "Kaplan-Meier Curve for Colon Cancer Recurrence \nby Treatment", 
           xlab = "Time in Days", 
           risk.table = T, risk.table.height = 0.25, risk.table.fontsize = 4, 
           tables.theme = theme_cleantable())
```

# 3. Log-Rank Test
```{r}
coxph(Surv(time, status) ~ rx, data = colon.recur)
summary(coxph(Surv(time, status) ~ rx, data = colon.recur))
```

# 4. Cox PH Model 
## 4.1 Model Selection
```{r}
# full model
r.model.full <- coxph(Surv(time, status) ~ ., data = colon.recur)

# stepwise selection with AIC criterion
r.model.aic <- step(r.model.full, direction = "both", k = 2)

anova(r.model.aic)
```

Stepwise selection method selected model:
Surv(time, status) ~ differ + rx + extent + surg + nodes

remove sex and obstruct

## 4.2 Model Diagnostic 
### 4.2.1 Check proportionality of hazard ratios
**Log of Negative Log of Estimated Survival Function**
```{r}
r.differ.fit <- survfit(Surv(time, status) ~ differ, data = colon.recur)
cll.differ = ggsurvplot(r.differ.fit, conf.int = F, font.x.size = 12, font.y.size = 12, font.legend.size = 9,
           fun = "cloglog",
           xlim = c(20, 4000),
           xlab = "Time (Days)",
           legend.lab = c("1-well","2-moderate","3-poor"),
           legend.title = 'Differ',
           title = "Log of Negative Log of Estimated Survival Function \nfor Colon Cancer Recurrence by Differ")
```

```{r}
r.extent.fit <- survfit(Surv(time, status) ~ extent, data = colon.recur)
cll.extent = ggsurvplot(r.extent.fit, conf.int = F, font.x.size = 12, font.y.size = 12, font.legend.size = 9,
           fun = "cloglog",
           xlim = c(20, 4000),
           xlab = "Time (Days)",
           legend.lab = c("1-submucosa","2-muscle","3-serosa", "4-contiguous structures"),
           legend.title = 'Extent',
           title = "Log of Negative Log of Estimated Survival Function \nfor Colon Cancer Recurrence by Extent")
```

```{r}
r.surg.fit <- survfit(Surv(time, status) ~ surg, data = colon.recur)
cll.surg = ggsurvplot(r.surg.fit, conf.int = F, font.x.size = 12, font.y.size = 12, font.legend.size = 9,
           fun = "cloglog",
           xlim = c(20, 4000),
           xlab = "Time (Days)",
           legend.lab = c("0-short","1-long"),
           legend.title = 'Surg',
           title = "Log of Negative Log of Estimated Survival Function \nfor Colon Cancer Recurrence by Surg")
```


```{r}
splots <- list()
splots[[1]] <- cll.differ
splots[[2]] <- cll.extent
splots[[3]] <- cll.surg


cloglog_plot = arrange_ggsurvplots(splots, print = FALSE, ncol = 2, nrow = 2)
ggsave(cloglog_plot,file = "r.C-log-log.pdf",width = 12,height = 15)
```

**Schoenfeld residuals**
```{r}
r.residual.fit = coxph(Surv(time, status) ~ differ + rx + extent + surg + nodes, data = colon.recur)
r.residual = cox.zph(coxph(Surv(time, status) ~ differ + rx + extent + surg + nodes, data = colon.recur))
r.residual


#residual plot for covariate that violate ph assumption
plot(r.residual[1], ylim = c(-2, 2))
abline(a = 0, b = 0, col = 2)
abline(h = r.residual.fit$coef[1], col = 3, lwd = 2, lty = 2)


plot(r.residual[5], ylim = c(-2, 2))
abline(a = 0, b = 0, col = 2)
abline(h = r.residual.fit$coef[8], col = 3, lwd = 2, lty = 2)


residual_plot = ggcoxzph(r.residual, font.main = 10, font.x = 10, font.y = 10, font.tickslab = 8,
         point.alpha = 0.5, point.col = "grey25")

ggsave("r.schoenfeld residual_plots.pdf", arrangeGrob(grobs = residual_plot))
```
differ violates ph assumption

## 4.3 Modification for Violation of PH Assumption
```{r}
#add interaction of covariate with function of time
r.model.inter = coxph(Surv(time, status) ~ rx + extent + surg + differ + tt(differ) + nodes, data = colon.recur, tt = function(x,t, ...) x*log(t))
anova(r.model.inter)
summary(r.model.inter)
```
differ has interaction with time


## 4.4 Final model
```{r}
r.model.final = coxph(Surv(time, status) ~ rx + extent + surg + differ + tt(differ) + nodes, data = colon.recur, tt = function(x,t, ...) x*log(t))
summary(r.model.inter)
```
